---
title: "Untitled"
author: "Chenyan Feng"
date: "2023-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1)	Develop Logistic Regression, LDA, QDA and KNN based survival prediction models using Pclass, Sex, Age, SibSp, Parch, and Embarked as predictor variables.  Note that some of these variables may need to be defined as categorical (factors in R).  Also, Age has lot of missing values. The missing values may need to be imputed (e.g., mean) for using this variable. Try few values of k in KNN to determine suitable value for K. Compare and interpret True Positive (TP) and False Positive (FP) of the different models using test data.


```{r}
# Load data
train_data = read.csv("ProjectTrain.csv")
test_data = read.csv("ProjectTest.csv")

# Convert Sex and Embarked to factors
train_data$Sex = factor(train_data$Sex)
train_data$Embarked = factor(train_data$Embarked)
test_data$Sex = factor(test_data$Sex)
test_data$Embarked = factor(test_data$Embarked)

# Impute missing values of Age with mean
train_data$Age[is.na(train_data$Age)] = mean(train_data$Age, na.rm = TRUE)
test_data$Age[is.na(test_data$Age)] = mean(train_data$Age, na.rm = TRUE)

# Calculate the contingency table and chi-square statistic for Sex and Embarked
table_SE = table(train_data$Sex, train_data$Embarked)
chisq_SE = chisq.test(table_SE)$statistic
# Calculate the correlation matrix for numeric variables
predictors <- train_data[, c("Pclass", "Age", "SibSp", "Parch")]
cor_matrix <- cor(predictors)
# Print the results
print("Contingency table for Sex and Embarked:")
print(table_SE)
print(paste0("Chi-square statistic for Sex and Embarked: ", chisq_SE))
print("Correlation matrix for numeric predictors:")
print(cor_matrix)


#p1
# Logistic regression model
log_model = glm(Survived ~ Pclass + Sex + Age + SibSp + Parch + Embarked, 
                 data = train_data, 
                 family = "binomial")
log_predictions = predict(log_model, newdata = test_data, type = "response")
log_predictions = ifelse(log_predictions > 0.5, 1, 0)
log_cm = table(log_predictions, test_data$Survived)
log_accuracy = sum(diag(log_cm)) / sum(log_cm)
cat(paste("Logistic regression accuracy:", log_accuracy, "\n"))

# LDA model
library(MASS)
lda_model <- lda(Survived ~ Pclass + Sex + Age + SibSp + Parch + Embarked, 
                 data = train_data)
lda_predictions <- predict(lda_model, newdata = test_data)
lda_cm <- table(lda_predictions$class, test_data$Survived)
lda_accuracy <- sum(diag(lda_cm)) / sum(lda_cm)
cat(paste("LDA accuracy:", lda_accuracy, "\n"))

# QDA model with regularization
qda_model <- qda(Survived ~ Pclass + Sex + Age + SibSp + Parch, 
                     data = train_data)
qda_predictions_reg <- predict(qda_model, newdata = test_data)
qda_cm_reg <- table(qda_predictions_reg$class, test_data$Survived)
qda_accuracy_reg <- sum(diag(qda_cm_reg)) / sum(qda_cm_reg)
cat(paste("QDA with regularization accuracy:", qda_accuracy_reg, "\n"))

# Convert Sex and Embarked to factors
train_data$Sex <- factor(train_data$Sex)
train_data$Embarked <- factor(train_data$Embarked)
test_data$Sex <- factor(test_data$Sex)
test_data$Embarked <- factor(test_data$Embarked)
# Encode factors as numeric values
train_data$Sex <- as.numeric(train_data$Sex) - 1
train_data$Embarked <- as.numeric(train_data$Embarked)
test_data$Sex <- as.numeric(test_data$Sex) - 1
test_data$Embarked <- as.numeric(test_data$Embarked)
# Fit KNN model
library(class)
set.seed(123)
# KNN model with k=5
knn_model <- knn(train = train_data[, c("Pclass", "Sex", "Age", "SibSp", "Parch", "Embarked")], 
                 test = test_data[, c("Pclass", "Sex", "Age", "SibSp", "Parch", "Embarked")], 
                 cl = train_data$Survived, 
                 k = 5)
knn_cm <- table(knn_model, test_data$Survived)
knn_accuracy <- sum(diag(knn_cm)) / sum(knn_cm)
cat(paste("KNN accuracy with k = 5:", knn_accuracy, "\n")) # Accuracy = 0.738

# KNN model with k=3
knn_model <- knn(train = train_data[, c("Pclass", "Sex", "Age", "SibSp", "Parch", "Embarked")], 
                 test = test_data[, c("Pclass", "Sex", "Age", "SibSp", "Parch", "Embarked")], 
                 cl = train_data$Survived, 
                 k = 3)
knn_cm <- table(knn_model, test_data$Survived)
knn_accuracy <- sum(diag(knn_cm)) / sum(knn_cm)
cat(paste("KNN accuracy with k = 3:", knn_accuracy, "\n")) # Accuracy = 0.753

# KNN model with k=7
knn_model <- knn(train = train_data[, c("Pclass", "Sex", "Age", "SibSp", "Parch", "Embarked")], 
                 test = test_data[, c("Pclass", "Sex", "Age", "SibSp", "Parch", "Embarked")], 
                 cl = train_data$Survived, 
                 k = 7)
knn_cm <- table(knn_model, test_data$Survived)
knn_accuracy <- sum(diag(knn_cm)) / sum(knn_cm)
cat(paste("KNN accuracy with k = 7:", knn_accuracy, "\n")) # Accuracy = 0.73

log_cm 
lda_cm
qda_cm_reg
knn_cm
```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
